Augmentation of a Model Historical Harpsichord Mechanism for Haptic Enabled Interaction

### 2025-01-20 src/article.tex

With the advent of digital musical instruments in the 20th century there has
been a decoupling between interface and sound generation when playing.
Consequently, aspects of the sensory experience of playing have been lost,
namely the sensation of vibration and mechanical impedance that a traditional
musical instrument provides.

The project aims to restore the tactile and haptic response of historical
harpsichords for an exhibition at the Museo di San
Colombano\footnote{https://genusbononiae.it/en/san-colombano-tagliavini-collection/}
in Bologna. San Colombano is home to the Tagliavini Collection, a collection of
approximately 70 historical keyboard instruments in working order
\cite{Tagliavini2007} gifted to the museum by the Museum.\\
from Luigi Ferdinando Tagliavini in 2010 \cite{SanColombano2010,Carlino2010}.

Some items in the Tagliavini collection are no longer in playing condition or
are too fragile to continue to be available to museum visitors. One of those
fragile instruments is a 1547 Harpsichord in the Italian style by Alessandro
Trasuntino \cite{Wraight2024}\footnote{examples of similar instruments by
Trasuntino available online in the Royal College of Music Collection (1531:
\url{https://museumcollections.rcm.ac.uk/collection/Details/collect/58}) and the
Musée de la musique collection (1538:
\url{https://carmentis.kmkg-mrah.be/eMP/eMuseumPlus?service=ExternalInterface&objectId=106088})
}, which was used as the basis for an interactive exhibition organised by San
Colombano.

The project was also open-sourced to allow for others to easily recreate and
iterate on the designs.

The "Haptic Key" project by Timmermans et al. \cite{Timmermans2020} began with
the model of a piano key. A 3-key cross-sectional model of a harpsichord
mechanism \ref{fig:3key} was provided to the NEMUS project by San Colombano as a
test bed for sensors and as proof of concept for a larger scale interface. Each
key of the model has 2 sets of strings that are under tension but not at any
specific musical pitch. The strings are there to provide the same resistance to
the jack plectra as would be found on an instrument. The mechanism is in the
Italian style of the XVI-XIX century with 2 jacks per key and their
corresponding strings tuned in unison \footnote{``8'-8' È la disposizione normale
dei clavicembali italiani'' \cite{Tagliavini1987}}.

A suitable sensing system was found to reliably detect when a jack had plucked a
string independently and then send a corresponding MIDI message to a connected
audio engine. A full scale 49-key scale model was commissioned with
and then built by luthier Roberto Livi in the same style as the 1547 Trasuntino
\ref{fig:teaser}. The sensor system was installed into the full-scale model and
then set up for exhibition at San Colombano. Visitors are invited to play on the
interface, which currently triggers samples on a connected sound engine. The
paper looks at the design decisions, processes, and considerations that need to
be made when undertaking a project of this kind.

@@

Overview of fabrication methods, techniques that can be employed.
Finally a discussion of the limitations and areas in which the core
project can be improved and developed further.

In this context, a `traditional' instrument is one whose interface for
generating sound and the sound generating mechanism are fundamentally
coupled. A piano requires a key press which triggers a hammer, which
strikes a string which vibrates a body and air cavity. This system is
tied together and cannot be separated. A synthesizer, be it a theremin,
a control voltage synthesizer, a MIDI controller, these are all
instruments whose interface are decoupled from their sound generation.
This decoupling presents a problem that then needs to be addressed. How
do we re-introduce this dimension to the experience of playing a musical
instrument? How do we, and is it possible to, parameterise and preserve
this experience? This is central problem that is trying to be solved in
\cite{Nichols2002, Timmermans2020, McAlpine2014, Baldwin2016}.

This is particularly pertinent as it relates to the conservation of musical
instruments, as they represent both historic objects, but also tools for
creating music. Historical musical instruments present an interesting
problem for cultural heritage as they are both a physical object, but
also a tool for making music. This begs the question, can we use haptics
to preserve experiences of the past? The proposed project intends to
develop further the methodologies implemented in musical haptics
projects \cite{MusicalHaptics2018} such as \cite{Timmermans2020}.

Systems exist already for the conversion of piano keyboard. The Yamaha
disklavier, the Don Buchla designed MOOG Piano Bar, Bosendorfer, PNOscan
by QRS focus on reproduction. These same systems could be applied to a
harpsichord, but the limitation is a single data stream per key.

\subsection{Motivations}\label{motivations}

The project was carried out in collaboration with San Colombano. The
results of the project would be presented as an interactive exhibition
for the public.

\begin{itemize}
\item
  existing sysetms

  \begin{itemize}
  \item
    Yamaha Disklavier used as a basis for exploration into haptics and
    the piano keyabord
    \cite{MusicalHaptics2018_04, MusicalHaptics2018_05,
    MusicalHaptics2018_13}
  \item
    Bosendorfer: LED photo transistor pair \cite{Moog1990} for obtaining
    performance data \cite{MusicalHaptics2018_05}

    \begin{itemize}
    \item
      290SE
    \item
      CEUS
    \end{itemize}
  \item
    Don Buchla / Moog Piano Bar
  \item
    PNOscan MIDI 9 QRS Music as MIDI Piano system \cite{McPherson2013}

    \begin{itemize}
    \item
      PNOmation
    \end{itemize}
  \item
    Piano Disc
  \end{itemize}
\end{itemize}

\section{Related Work}\label{related-work}

Methodologies for increasing visitor engagmemnet or

Museum staff want visitors to be engaged but limits in staffing and
funding limit how vistors can interact with collections
\cite{Templeton2018,
McAlpine2014}. For musical instrument museums there is an additional
difficulty in the balance of keeping historical instruments playable
\cite{McAlpine2014}. The fragility and decay of musical instruments
means that there is an inveitable point where instruments will no longer
be in a playable condition \cite{McAlpine2014, Fritz2017}

McAlpine outlines a similar situation to the Tagliavini collection in
his work with the Benton Fletcher Collection \cite{McAlpine2014}. There
was a stipulation by Benton Fletcher that the instruments remain
available to play when they were gifted to National Trust Fenton House.
A custom MIDI interface arranged in a two manual harpsichord style which
triggered samples of the instruments in the collection using an
appropraite recording strategy for each. One problem highlighted in user
tests by McAlpine is that the weighted keys did no provide ``an
experiential sense of interacting with a historical keyboard''
\cite{McAlpine2014}. The MIDI interface created from commercially
available weighted keys. McAlpine posits a haptic keyboard augemented
with ``actuators to provide positionally-sensitive real-time force
feedback at point-of-contact'' in a similar approach as the piano
mechanism by Gillespie \cite{Gillespie1996}.

Previous NIME project ``Tromba Moderna'' \cite{Baldwin2016} looked to
avoid the complex engineering problem by simply recreating the tromba
marina and augmenting it. In the case of the Tromba Modern, a piezo
transducer was connected to a sound synthesis engine connected to a
driver inside the instrument to simulate the vibration that would be
expected of an historical tromba marina.

The original design was to take a similar approach to the project by
Timmermans \cite{Timmermans2020}, which was an extension of the system
by Gillespie \cite{Gillespie1996}. The process laid out in Timmermans
project was to begin with a single key model of the mechanism. We began
with a 3-key Model Harpsichord Mechanism by Graziano Bandini
\textbackslash figure\{\}. Before we continued with augementing a model
with actuators it was decided that it we should first validate whether a
recreation of the harpsichord mechanism would suffice in allowing
visitors to suspend disbelief.

Benefits of putting effort into the maintenance of one interface with a
transferable sensor system

An additonal factor not faced in the Tromba Moderna project is the
problem of scale in managing 98 rather than a single data stream of
data.

Material costs mitigated by creating a scaled version of the sound
board, providing only the length of string necessary under comparable
tension and thus resistance to the jack quill.

Finally, a further extension to the projects above was a commmitement to
open sourcing all aspects of the project, hardware, software, and data.

@@

@@

The approach taken was to recreate the interface of these instruments
not for acoustics purposes be purely for the mechanical resoponse. As
exhibition for the general public in a musical instruments there were
aesthetic and functional constraints placed on the design of the
interface.

@@

\subsubsection{Anatomy of Harpsichord
mechanics}\label{anatomy-of-harpsichord-mechanics}

\begin{itemize}
\item
  simple
\item
  aesthetics
\item
  split workload
\item
  parts of harpsichord

  \begin{itemize}
  \item
    jack

    \begin{itemize}
    \item
      quill
    \item
      jack body
    \item
      pivot
    \item
      tongue
    \end{itemize}
  \item
    key
  \item
    body
  \end{itemize}
\end{itemize}

\subsubsection{Keyboard Construction}\label{keyboard-construction}

The desicion was made to design the elctronic sna dthe model keyboard in
tandem. Allow for quicker development time and iteration on the idea.
Likely the first implementation would need re-designed and redesiginng
the electronics to fit the form of the keyboard was more cost effective
than building or carrying out major modifications to the model.

@@

Discussion of spatial constraints and affects on electyronics design and
installation is covered in \textbackslash section\{setup\}

The rear chamber was desigbned to house comonents for audio processing,
which is discussed further in \textbackslash section

\footnote{https://github.com/Nemus-Project/haptic-harpsichord-nime-2025/issues/Nemus-Project/haptic-harpsichord-nime-2025/images/disegno.pdf}

For the initial prototyping stage, modifying an existing harpsichord was
considered though ultimately discarded. An existing instrument would
have provided a test for scaling the electronics, but internal
measurments and layout would have been too different. A harpsichord that
was economic enough to be deconstructed would also have been of a more
modern mechanical design and not of the design that was being recreated
with the model. Prototyping stage would likely be invasive and damage or
heavy modification likely. As such it was not a good use of funds as a
modern instrument fit for destruction did not offer benefits that the
small model with easy internal access did not already provide.

\subsection{Building Process}\label{building-process}

PARAGRAPH FROM ROBERTO

@@

\section{Related Work and Motivation}\label{related-work-and-motivation}

\begin{itemize}
\item
  \cite{Timmermans2020}
\item
  \cite{McPherson2013} \cite{McPherson2019}
\item
  \cite{Mudd2013}
\item
  \cite{Fritz2017}
\end{itemize}

@@

Using a similalr apporach to @HapticKey the first step was to fabricate
or procure a model of a single working key. San Colombano provided a
model 3-key of a harposchord mechanism by Graziano Bandini
\textbackslash FIGURE. The 3-key model was used a test-bed for potential
sensors while designs were drawn for a full-scale model. The
cross-sectional nature of the model meant it was easy to fit and remove
sensors. Small size meant that it would also be easy to transfer between
workshops.

3-key model also had a sliding jack rail which allowed for easy
disengagement of the physical mechanical response, which allowed for
some preliminary tests on the presence and absence of mechanism.

@@

A number of sensors were tried before the final system was decided upon.
An inertial measurment unit (IMU) was mounted to a single key which
contained a 6 degrees of freedome acceleromoetr and gyroscope. There was
a consistent set of conditions to trigger sound, but this was not easily
differntiable when other keys were played. The fitting time and cost of
a single IMU could b reduced when purchased in bulk but was far greater
than other systems considered.

A reed sensor and a hall sensor were mounted and magnets embedded within
the jacks. Adjusting the hieight of the sensor provided a reliable means
of detecting a threshold, inparticular the hall sensors which could be
partially calibrarted physically and also in softwrae. The time taken to
calibarte a single key was time consuming. Ideally each sensor would be
placed on a thread for adjustment, but given the constraints on the
internal space it was decided any adjustment of heights would likely
mean an unnacceptable length of time to install and calibrate. Embedding
magnets into the jacks was also did not satisfy the criteria of being
scalable and non-invasive.

@@

Finally, force sensitive resistors placed under the jack of each key was
considered but was never implemented. The sensors were to be placed
under each jack either above or below the padding on which the bottom of
the jack rests. Further discussion of this apporach is in
\textbackslash section.

@@

\begin{itemize}
\item
  previous use
\item
  problem of what to sense
\item
  satisfied both criteria and design restrictions
\item
  usage

  \begin{itemize}
  \item
    Sweet spot is around 6mm distance
  \end{itemize}
\end{itemize}

\subsubsection{sensor surface}\label{sensor-surface}

\begin{itemize}
\item
  linear gradient
\item
  on Jack
\item
  printed and cut on vinyl {[}find name{]}
\item
  cut to travel size
\item
  then to jack size
\item
  jacks varnished on side
\item
  alternatives

  \begin{itemize}
  \item
    CNC plywood and mdf

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \item
    laser cut plywood and mdf

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \item
    3D printed jack body

    \begin{itemize}
    \item
      problems
    \end{itemize}
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Nano form factor small
\item
  allows for easy changes in chipset
\item
  ``Native USB''
  \footnote{https://learn.adafruit.com/dude-where-s-my-com-port/native-usb-boards}
\item
  BLE Model

  \begin{itemize}
  \item
    Performative
  \item
    BLE capabilities for MIDI BLE transmission
  \item
    All analog pins available
  \item
    available at the time
  \item
    12-bit DAC
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  On board available, but volatile
\item
  retained across firmware changes
\item
  small,
\item
  could have been sd card or anything else
\item
  available
\item
  interfaced via rotary encoder
\end{itemize}

@@

\begin{itemize}
\item
  locating keys
\item
  displaying key state
\end{itemize}

@@

The multiplexer reduced 7 signals to a single pin. Three pins are
required pins for addressing the muliplexewrs and all PCBs are addressed
simulatenously.

\begin{itemize}
\item
  limited ADC channels
\item
  ADC channels already multiplexed
\item
  simple
\item
  8 channel
\item
  THT
\item
  7 into 49

  \begin{itemize}
  \item
    7 PCBS
  \item
    each with 7 sensors
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  EAGLE PCBs designed using EAGLE. Worth noting that AutoCAD has
  deprecated EAGLE and the programme will no longer be supported from
  2026 Project files will need to be transitioned to another standard,
  likely the KiCAD programme.
\end{itemize}

@@

  \begin{itemize}
  \item
    striping all adc lines with FFC
  \item
    solder pads to select channel
  \end{itemize}
\end{itemize}

@@

The first version of the board esclusively used 2.54mm pitch headers.
The reasons for this were flexibility when creating cable looms and to
leave open the possiility of rewuring the PCB.

This was the case for the LEDs as they initially wer to be powered on
5V. It was discovered that the 3v regulator of the nano was sufficient
to power all LEDs at a lower brightness while reducing current draw.
Since brightness was not importnat.

@@

\begin{itemize}
\item
  https://www.thingiverse.com/thing:1646016
\item
  https://www.thingiverse.com/thing:2886460
\item
  Supports

  \begin{itemize}
  \item
    Push fit
  \item
    adjustable for best average placement
  \end{itemize}
\item
  Blinkers

  \begin{itemize}
  \item
    reduce noise from other senors
  \end{itemize}
\item
  mounts

  \begin{itemize}
  \item
    MacMini
  \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item
  Arduino platform

  \begin{itemize}
  \item
    library support and portability
  \item
    dependancy management
  \item
    wide spread use
  \end{itemize}
\item
  core algorithm

>>>>

Technological advances have transformed how museums document, present and interpret their collections, fostering immersive experiences through tools such as laser scanning, 3D printing, and virtual reality \cite{allard2005use,Wachowiak01082009,RCM_2024_3D,Kuzminsky_LaserScan_2012,Schaich_3D_2007}. These technologies place an emphasis on experiential authenticity, enabling encounters that evoke the sensory, emotional, and intellectual essence of the past \cite{trant_Auth_1999}. However, as Pine and Gilmore note \cite{pinegilmore_2007}, achieving authenticity requires museums to navigate the delicate balance between preservation and meaningful engagement—a challenge that is particularly evident in the case of historical musical instrument collections \cite{McAlpine2014}.

Musical instruments represent a distinctive fusion of form, function, and history. Their cultural value extends beyond their visual appeal to include the tactile and auditory dimensions of use \cite{Fritz2017}. Yet, preservation concerns often limit direct interaction, reducing these artefacts to static displays. This ``red velvet cord'' approach, as theorised by McAlpine \cite{McAlpine2014}, protects fragile mechanisms but diminishes the instruments’ functional identity, disconnecting visitors from the full richness of their historical and cultural context.

The Tagliavini Collection in Bologna \cite{Tagliavini2007}, renowned for its historical keyboard instruments, exemplifies this dilemma. To address it, this project introduces a haptic keyboard, a digital interface designed to bridge the gap between preservation and accessibility. The project builds on Andrew McPherson’s earlier work on "Portable Measurement and Mapping of Continuous Piano Gesture" and aligns with principles outlined in The O in NIME, which emphasises the importance of extending the lifecycle of existing NIMEs (New Interfaces for Musical Expression). By reusing, updating, and complementing previous NIMEs, the project fosters sustainability and ensures their relevance beyond their original context.

The haptic keyboard combines high-definition sound sampling, touch-sensitive technology, and ergonomic design, recreating the sound and tactile qualities of the Tagliavini instruments without compromising their physical integrity.

The article is organised as follows: TBC

% With the advent of digital musical instruments in the 20th century there has
% been a decoupling between interface and sound generation when playing.
% Consequently, aspects of the sensory experience of playing have been lost,
% namely the sensation of vibration and mechanical impedance that a traditional
% musical instrument provides.

% The project aims to restore the tactile and haptic response of historical
% harpsichords for an exhibition at the Museo di San
% Colombano\footnote{https://genusbononiae.it/en/san-colombano-tagliavini-collection/}
% in Bologna. San Colombano is home to the Tagliavini Collection, a collection of
% approximately 70 historical keyboard instruments in working order
% \cite{Tagliavini2007} gifted to the museum by Luigi Ferdinando
% Tagliavini in 2010 \cite{SanColombano2010,Carlino2010}.

% Some items in the Tagliavini collection are no longer in playing condition or
% are too fragile to continue to be available to museum visitors. One of those
% fragile instruments is a 1547 Harpsichord in the Italian style by Alessandro
% Trasuntino \cite{Wraight2024}\footnote{examples of similar instruments by
% Trasuntino available online in the Royal College of Music Collection (1531:
% \url{https://museumcollections.rcm.ac.uk/collection/Details/collect/58}) and the
% Musée de la musique collection (1538:
% \url{https://carmentis.kmkg-mrah.be/eMP/eMuseumPlus?service=ExternalInterface&objectId=106088})
% }, which was used as the basis for an interactive exhibition arranged by San
% Colombano. Part of the exhibit is a keyboard that replicates the mechanism of
% the Trasuntino that aims to provide a similar tactile playing experience to that
% of the original.

% The project was also open-sourced to allow for others to easily recreate and
% iterate on the designs.

% The "Haptic Key" project by Timmermans et al. \cite{Timmermans2020} began with
% the model of a piano key. A 3-key cross-sectional model of a harpsichord
% mechanism \ref{fig:3key} was provided to the NEMUS project by San Colombano as a
% test bed for sensors and as proof of concept for a larger scale interface. Each
% key of the model has 2 sets of strings that are under tension but not at any
% specific musical pitch. The strings are there to provide the same resistance to
% the jack plectra as would be found on an instrument. The mechanism is in the
% Italian style of the XVI-XIX century with 2 jacks per key and their
% corresponding strings tuned in unison \footnote{``8'-8' È la disposizione normale
% dei clavicembali italiani'' \cite{Tagliavini1987}}.

% A suitable sensing system was found to reliably detect when a jack had plucked a
% string independently and then send a corresponding MIDI message to a connected
% audio engine. A full scale 49-key scale model was commissioned with and then
% built by the luthier \anon{Roberto Livi} in the same style as the 1547 Trasuntino
% \ref{fig:teaser}. The sensor system was installed into the full-scale model and
% then set up for exhibition at San Colombano. Visitors are invited to play on the
% interface, which currently triggers samples on a connected sound engine. The
% paper looks at the design decisions, processes, and considerations that need to
% be made when undertaking a project of this kind.

@@

% Overview of fabrication methods, techniques that can be employed.
% Finally a discussion of the limitations and areas in which the core
% project can be improved and developed further.

% In this context, a `traditional' instrument is one whose interface for
% generating sound and the sound generating mechanism are fundamentally
% coupled. A piano requires a key press which triggers a hammer, which
% strikes a string which vibrates a body and air cavity. This system is
% tied together and cannot be separated. A synthesizer, be it a theremin,
% a control voltage synthesizer, a MIDI controller, these are all
% instruments whose interface are decoupled from their sound generation.
% This decoupling presents a problem that then needs to be addressed. How
% do we re-introduce this dimension to the experience of playing a musical
% instrument? How do we, and is it possible to, parameterise and preserve
% this experience? This is central problem that is trying to be solved in
% \cite{Nichols2002, Timmermans2020, McAlpine2014, Baldwin2016}.

% This is particularly pertinent as it relates to the conservation of musical
% instruments, as they represent both historic objects, but also tools for
% creating music. Historical musical instruments present an interesting
% problem for cultural heritage as they are both a physical object, but
% also a tool for making music. This begs the question, can we use haptics
% to preserve experiences of the past? The proposed project intends to
% develop further the methodologies implemented in musical haptics
% projects \cite{MusicalHaptics2018} such as \cite{Timmermans2020}.

% Systems exist already for the conversion of piano keyboard. The Yamaha
% disklavier, the Don Buchla designed MOOG Piano Bar, Bosendorfer, PNOscan
% by QRS focus on reproduction. These same systems could be applied to a
% harpsichord, but the limitation is a single data stream per key.

% \subsection{Motivations}\label{motivations}

% \begin{itemize}
% \item
%   existing sysetms
%   \begin{itemize}
%   \item
%     Yamaha Disklavier used as a basis for exploration into haptics and
%     the piano keyabord
%     \cite{MusicalHaptics2018_04, MusicalHaptics2018_05,
%     MusicalHaptics2018_13}
%   \item
%     Bosendorfer: LED photo transistor pair \cite{Moog1990} for obtaining
%     performance data \cite{MusicalHaptics2018_05}

%     \begin{itemize}
%     \item
%       290SE
%     \item
%       CEUS
%     \end{itemize}
%   \item
%     Don Buchla / Moog Piano Bar
%   \item
%     PNOscan MIDI 9 QRS Music as MIDI Piano system \cite{McPherson2013}

%     \begin{itemize}
%     \item
%       PNOmation
%     \end{itemize}
%   \item
%     Piano Disc
%   \end{itemize}
% \end{itemize}

\section{Related Work and Motivations}\label{related-work}

% These systems mentioned could be adapted for a harpsichord, but they would not
% easily capture the separate plucks from both jacks per key. We decided to create
% our own system and focus on detecting the vertical displacement of the jacks
% directly, rather than the position of the key.

% Methodologies for increasing visitor engagmemnet or

% Museum staff want visitors to be engaged but limits in staffing and
% funding limit how vistors can interact with collections
% \cite{Templeton2018,
% McAlpine2014}. For musical instrument museums there is an additional
% difficulty in the balance of keeping historical instruments playable
% \cite{McAlpine2014}. The fragility and decay of musical instruments
% means that there is an inveitable point where instruments will no longer
% be in a playable condition \cite{McAlpine2014, Fritz2017}

% McAlpine outlines a similar situation to the Tagliavini collection in
% his work with the Benton Fletcher Collection \cite{McAlpine2014}. There
% was a stipulation by Benton Fletcher that the instruments remain
% available to play when they were gifted to National Trust Fenton House.
% A custom MIDI interface arranged in a two manual harpsichord style which
% triggered samples of the instruments in the collection using an
% appropraite recording strategy for each. One problem highlighted in user
% tests by McAlpine is that the weighted keys did no provide ``an
% experiential sense of interacting with a historical keyboard''
% \cite{McAlpine2014}. The MIDI interface created from commercially
% available weighted keys. McAlpine posits a haptic keyboard augemented
% with ``actuators to provide positionally-sensitive real-time force
% feedback at point-of-contact'' in a similar approach as the piano
% mechanism by Gillespie \cite{Gillespie1996}.

% Previous NIME project ``Tromba Moderna'' \cite{Baldwin2016} looked to
% avoid the complex engineering problem by simply recreating the tromba
% marina and augmenting it. In the case of the Tromba Modern, a piezo
% transducer was connected to a sound synthesis engine connected to a
% driver inside the instrument to simulate the vibration that would be
% expected of an historical tromba marina.

% The original design was to take a similar approach to the project by
% Timmermans \cite{Timmermans2020}, which was an extension of the system
% by Gillespie \cite{Gillespie1996}. The process laid out in Timmermans
% project was to begin with a single key model of the mechanism. We began
% with a 3-key Model Harpsichord Mechanism by Graziano Bandini
% \textbackslash figure\{\}. Before we continued with augementing a model
% with actuators it was decided that it we should first validate whether a
% recreation of the harpsichord mechanism would suffice in allowing
% visitors to suspend disbelief.

% Benefits of putting effort into the maintenance of one interface with a
% transferable sensor system

% An additonal factor not faced in the Tromba Moderna project is the
% problem of scale in managing 98 rather than a single data stream of
% data.

% Material costs mitigated by creating a scaled version of the sound
% board, providing only the length of string necessary under comparable
% tension and thus resistance to the jack quill.

% Finally, a further extension to the projects above was a commmitement to
% open sourcing all aspects of the project, hardware, software, and data.

% There are commercial systems for augmenting piano keyboards generating MIDI data.

% An additional goal of the project was to design an interface that could be used to research
% the influence the mechanism and it's resistances have on a player's performance.
% For example, the Yamaha Disklavier series of pianos have an embedded sensor
% system that allows for the instruments to operate in a "silent mode", but still
% output MIDI data when the keys are struck. This has been a rich avenue for
% exploring the affect of vibration on performance,
% \cite{MusicalHaptics2018_04,MusicalHaptics2018_13} perception in instrument
% quality \cite{MusicalHaptics2018_05}.

% As well as a trying to offer a more engaging experience to museum attendees, the
% interface was also designed as a potential probe for further research into the
% idiosyncrasies of the harpsichord and the influence the mechanism has on
% performance.

Museums often face significant challenges in engaging visitors due to limitations in staffing and funding, which restrict the ways in which visitors can interact with collections \cite{Templeton2018, McAlpine2014}. For musical instrument museums, these challenges are compounded by the difficulty of preserving historical instruments in a playable condition \cite{McAlpine2014}. The inherent fragility and gradual decay of such instruments inevitably result in a point where they can no longer be played. A significant cultural change has taken place in recent decades, shifting the focus from the playability of the originals to their conservation. Karp \cite{Karp1979,Karp1985} advocates for a deeper understanding of musical instruments so that enough knowledge is generated to make them as ``copyable'' as possible.

McAlpine discusses a case similar to the Tagliavini collection in his examination of the Benton Fletcher Collection at National Trust Fenton House \cite{McAlpine2014}. When these instruments were donated, Benton Fletcher stipulated that they remain playable and should continue to be maintained for tuition and public performance. A large sampling campaign was conducted, and a custom MIDI interface was designed to fulfil this requirement while preserving the original instruments' integrity. The MIDI keyboard, comprising two commercially available keyboards mimicking the two-manual harpsichord layout, were used by visitors to trigger the instrument samples recorded with tailored strategies for each. However, user tests identified a significant limitation: the commercially available weighted keys failed to provide an authentic sense of interacting with a historical keyboard \cite{McAlpine2014}. To address this, McAlpine proposed -- though did not follow up to -- a haptic keyboard enhanced with actuators capable of delivering positionally sensitive, real-time force feedback at the point of contact, inspired by Gillespie’s work on piano mechanisms \cite{Gillespie1996}.

The ``Tromba Moderna'' project \cite{Baldwin2016}, a previous NIME initiative, approached this issue differently by recreating and augmenting a replica of a historical tromba marina. A piezo transducer was connected to a sound synthesis engine and a driver within the instrument to simulate the expected vibrations of a historical tromba marina. This approach avoided the complex engineering challenges posed by fully replicating the original mechanisms.

This work inherits the same philosophy as the Tromba Moderna project. By augmenting a replica of a historical harpsichord keyboard using minimally invasive electronics, the project intends to offer a tool enhancing the fruition of the Tagliavini collection whilst retaining a form of continuity with historical musical instrument-building traditions.

Beyond enhancing the visitor experience, the interface serves as a research tool to explore the unique characteristics of the harpsichord and its impact on performance.

@@

The present project initially drew inspiration from Timmermans' extension of Gillespie’s system \cite{Timmermans2020, Gillespie1996}. Timmermans’ process began with a single-key model of the mechanism. Similarly, we started with a 3-key model harpsichord mechanism, crafted by Graziano Bandini. Before incorporating actuators into the model, we sought to validate whether a faithful recreation of the harpsichord mechanism could effectively enable visitors to suspend disbelief.

A critical benefit of focusing on the development of a single interface with a transferable sensor system is the potential for scalability and adaptability. However, managing data streams presents a significant challenge. Unlike the Tromba Moderna project, which dealt with a single data stream, our design must manage 98 simultaneous data streams. This complexity necessitated cost-effective solutions, such as scaling down the soundboard to include only the necessary string length under comparable tension and resistance to the jack quill.

Another distinguishing feature of this project is its commitment to open-sourcing all hardware, software, and data components. While commercial systems for augmenting piano keyboards and generating MIDI data exist, this project emphasizes accessibility and transparency.

An additional objective was to design an interface that supports research into how the harpsichord mechanism and its resistances influence a player’s performance. For example, the Yamaha Disklavier series incorporates embedded sensors enabling "silent mode" operation while outputting MIDI data when keys are struck. This has facilitated research into the effects of vibration on performance \cite{MusicalHaptics2018_04, MusicalHaptics2018_13} and the perception of instrument quality \cite{MusicalHaptics2018_05}.

@@

The approach taken was to recreate the interface of these instruments
not for acoustics purposes be purely for the mechanical response. As
exhibition for the general public in a musical instruments there were
aesthetic and functional constraints placed on the design of the
interface.

@@

\subsubsection{Keyboard Construction}\label{keyboard-construction}
% PARAGRAPH FROM ROBERTO
The construction of the keyboard required three months of work to be fully completed
and followed the same construction phases as a real harpsichord.

The model was constructed using the materials of the Italian harpsichord
tradition:

\begin{itemize}
\item
  Walnut wood for the wrestplank
\item
  Chestnut wood for the key levers
\item
  Boxwood and ebony for the key caps.
\item
  Cypress wood for the soundboard, a wood characterized by a
\item
  Beech wood, a brass spring and a seagull feather plectrum for the 98 jacks
\end{itemize}

The body of the case is based on a rectangular poplar frame that leaves space
for the installation of the electronic sensors.

Deviations from the design of the Trasuntino are the rectangular, rather than
logartithmic shape, removal of a bottom panel and ommision of a short octave,
instead spanning 4 octaves beginning on C.

The strings are Malcolm Rose yellow brass wire, which are secured to wrought
iron pins. The tension of each string was adjusted to provide resistance to
plucking close to that of a harpsichord. Three felt strips were woven between
the strings so as to cancel the vibration of the strings.

The result is an interface instrument that is placed between the mechanical
action of the keys and the generation of a synthetic sound. The touch of the
wooden keys, the friction of the natural feathers on the string, the noise of
the percussion at the base of the key together produce the sensation of being in
front of a real harpsichord.

% \subsubsection{Anatomy of Harpsichord
% mechanics}\label{anatomy-of-harpsichord-mechanics}

% \begin{itemize}
% \item
%   simple
% \item
%   aesthetics
% \item
%   split workload
% \item
%   parts of harpsichord

%   \begin{itemize}
%   \item
%     jack

%     \begin{itemize}
%     \item
%       quill
%     \item
%       jack body
%     \item
%       pivot
%     \item
%       tongue
%     \end{itemize}
%   \item
%     key
%   \item
%     body
%   \end{itemize}
% \end{itemize}

% The decision was made to design the electronics and the model keyboard simultaneously. 
% This meant development development time was quicker and iteration on the idea.
% Likely the first implementation would need re-designed and redesiginng
% the electronics to fit the form of the keyboard was more cost effective
% than building or carrying out major modifications to the model.

@@

% Discussion of spatial constraints and effects on electronics design and
% installation is covered in \textbackslash section\{setup\}

The rear chamber was designed to house components for audio processing,
which is discussed further in \textbackslash section

\begin{figure}  
  \centering
  \includegraphics[width=\linewidth]{images/49-key-bottom.jpeg} 
  \caption{Underside of the full model keyboard showing two chambers. Front chamber at the top and rear chamber at the bottom.} 
  \Description{} 
  \label{fig:49-key-bottom}
\end{figure}

% \footnote{https://github.com/Nemus-Project/haptic-harpsichord-nime-2025/issues/Nemus-Project/haptic-harpsichord-nime-2025/images/disegno.pdf}

For the initial prototyping stage, modifying an existing harpsichord was
considered though ultimately discarded. An existing instrument would
have provided a test for scaling the electronics, but internal
measurements and layout would have been too different. A harpsichord that
was economic enough to be deconstructed would also have been of a more
modern mechanical design and not of the design that was being recreated
with the model. Prototyping stage would likely be invasive and damage or
heavy modification likely. As such it was not a good use of funds as a
modern instrument fit for destruction did not offer benefits that the
small model with easy internal access did not already provide.

@@

% \section{Related Work and Motivation}\label{related-work-and-motivation}

% \begin{itemize}
% \item
%   \cite{Timmermans2020}
% \item
%   \cite{McPherson2013} \cite{McPherson2019}
% \item
%   \cite{Mudd2013}
% \item
%   \cite{Fritz2017}
% \end{itemize}

@@

% Using a similar apporach to @HapticKey the first step was to fabricate
% or procure a model of a single working key. San Colombano provided a
% model 3-key of a harposchord mechanism by Graziano Bandini
% \textbackslash FIGURE. The 3-key model was used a test-bed for potential
% sensors while designs were drawn for a full-scale model. The
% cross-sectional nature of the model meant it was easy to fit and remove
% sensors. Small size meant that it would also be easy to transfer between
% workshops.

% 3-key model also had a sliding jack rail which allowed for easy
% disengagement of the physical mechanical response, which allowed for
% some preliminary tests on the presence and absence of mechanism.

@@

A number of sensors were tried before the final system was decided upon.
An inertial measurement unit (IMU) was mounted to a single key which
contained a 6-degrees-of-freedom accelerometer and gyroscope. There was
a consistent set of conditions to trigger sound, but this was not easily
differentiable when other keys were played. The fitting time and cost of
a single IMU could b reduced when purchased in bulk but was far greater
than other systems considered.

A reed sensor and a hall sensor were mounted and magnets embedded within
the jacks. Adjusting the height of the sensor provided a reliable means
of detecting a threshold, in particular the hall sensors which could be
partially calibrated physically and also in software. The time taken to
calibrate a single key was time consuming. Ideally each sensor would be
placed on a thread for adjustment, but given the constraints on the
internal space it was decided any adjustment of heights would likely
mean an unacceptable length of time to install and calibrate. Embedding
magnets into the jacks was also did not satisfy the criteria of being
scalable and non-invasive.

@@

Finally, force sensitive resistors placed under the jack of each key was
considered but was never implemented. The sensors were to be placed
under each jack either above or below the padding on which the bottom of
the jack rests.

% Further discussion of this approach in \textbackslash section.

@@

@@

@@

@@

@@

The multiplexer reduced 7 signals to a single pin. Three pins are
required pins for addressing the multiplexers and all PCBs are addressed
simultaneously.

@@

EAGLE PCBs designed using EAGLE. Worth noting that AutoCAD has
deprecated EAGLE and the programme will no longer be supported from
2026 Project files will need to be transitioned to another standard,
likely the KiCAD programme.

@@

  \begin{itemize}
  \item
    striping all ADC lines with FFC
  \item
    solder pads to select channel
  \end{itemize}
\end{itemize}

@@

The first version of the board exclusively used 2.54mm pitch headers.
The reasons for this were flexibility when creating cable looms and to
leave open the possibility of rewiring the PCB.

This was the case for the LEDs as they initially were to be powered on
5V. It was discovered that the 3v regulator of the nano was sufficient
to power all LEDs at a lower brightness while reducing current draw.
Since brightness was not important.

@@

@@

% \subsection{Project structure}\label{project-structure}

% The project is separated into three repositories encapsulating
